---
title: "Clustering-Based Pairs Trading on S&P 500 Sector ETFs"
subtitle: "FIN 452 — Quantitative Trading Project"
author: "Soumer"
date: "`r Sys.Date()`" 
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: flatly
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
  echo: false
---






## 1. Executive Summary

 ![Strategy Mental Model](assets/mental_model.svg)
 
 
+ This project implements a two-layer clustering-based pairs trading strategy on 12 S&P 500 Select Sector SPDR ETFs. Layer 1 selects tradeable pairs monthly via k-means and hierarchical clustering on mean-reversion features (half-life, Hurst exponent, spread volatility). Layer 2 generates daily trade signals using a Kalman Filter that tracks dynamic hedge ratios and spread deviations. The strategy targets risk-adjusted outperformance of the SPY benchmark over the January 2025 -- March 2026 test period.

## 2. Universe & Data

```{r}
#| label: Libraries

library(tidyverse)
library(RTL)
library(xts)
library(TTR)
library(tidyquant)
library(pracma)
library(doParallel)
library(foreach)
library(parallel)
library(gt)
library(patchwork)
library(dlm)
library(PerformanceAnalytics)
```




```{r}
#| label: Parameters

train_start  <- as.Date("2000-06-01")
train_end    <- as.Date("2024-12-31")
test_start   <- as.Date("2025-01-01")
test_end    <- as.Date("2026-03-31")
window_roll  <- 252L
entry_zscore <- 1.5
max_innov_days <- 5L
innov_mult   <- 3.0
top_pairs  <- 10L
k_min        <- 2L
k_max        <- 12L
XLC_inception <- as.Date("2018-06-19")
XLRE_inception <- as.Date("2015-10-07")
```
### 2.1 Assets Considered

+ The universe consists of 11 sector ETFs plus SPY as both benchmark and tradeable asset. XLC (Communication Services) is only available from June 2018 and XLRE (Real Estate) from October 2015. Pairs involving these tickers are excluded before their inception dates.

```{r}
#| label: etfs-info-table
#| 
tickers <- c("SPY","XLB","XLC","XLE","XLF","XLI","XLK","XLP","XLRE","XLV","XLU","XLY")

ticker_info <- tibble::tibble(
  ticker = tickers,
  sector = c("S&P 500 Benchmark", "Materials", "Communication Services",
             "Energy", "Financials", "Industrials", "Technology",
             "Consumer Staples", "Real Estate", "Health Care",
             "Utilities", "Consumer Discretionary"),
  inception_note = c("", "", "From 2018-06-19", "", "", "", "",
                     "", "From 2015-10-07", "", "", "")
)

ticker_info %>% gt::gt() %>%
  gt::tab_header(title = "Asset Universe") %>%
  gt::cols_label(ticker = "Ticker", sector = "Sector",
                 inception_note = "Note")
```


```{r}
#| label: Data Download

dat_raw <- tidyquant::tq_get(tickers,
                             get = "stock.prices",
                             from = train_start,
                             to = test_end + 1) %>% 
  dplyr::select(date,symbol,adjusted) %>% 
  dplyr::rename(ticker = symbol, price = adjusted)

dat_wide <- dat_raw %>% 
  tidyr::pivot_wider(names_from = ticker,
                     values_from = price) %>% 
  dplyr::arrange(date)

spy_prices <- dat_raw %>%
  dplyr::filter(ticker == "SPY") %>%
  dplyr::select(date, price) %>%
  dplyr::mutate(spy_ret = price / dplyr::lag(price) - 1)


```

### 2.2 Pair Definitions

+ We enumerate all 66 unique pairs from the 12 assets. Each pair has a validity start date determined by the later inception date of its two legs.

```{r}
#| label: Pair Construction

pair_grid <- tidyr::expand_grid(
  leg1 = tickers,
  leg2 = tickers
) %>%
  dplyr::filter(leg1 < leg2) %>%
  dplyr::mutate(
    pair_id = paste0(leg1, "_", leg2),
    valid_from = dplyr::case_when(
      leg1 == "XLC" | leg2 == "XLC" ~ XLC_inception,
      leg1 == "XLRE" | leg2 == "XLRE" ~ XLRE_inception,
      TRUE ~ train_start
    )
  )
```


## 3. Layer 1: Monthly Pair Selection

### 3.1 Spread Feature Computation

+ For each of the 66 pairs, we compute three rolling features that characterize mean-reversion behavior. The spread is defined as `S_t = p1_t - beta_hat * p2_t` where `beta_hat` comes from an OLS regression over the rolling window_roll. Half-life measures speed of reversion, Hurst exponent measures persistence structure (H < 0.5 is mean-reverting), and spread volatility captures the magnitude of movements.

```{r}
#| label: Features-funcs

half_life <- function(spread) {
  
  n <- length(spread)
  ds <- diff(spread)
  s_lag <- spread[-n]
  
  fit <- lm(ds ~ s_lag)
  beta <- coef(fit)[2]
  -log(2) / beta
}


hurst_exp <- function(spread){
  
  pracma::hurstexp(spread,d = 20,display = F)$Hrs
}


spread_vol <- function(spread){
  
  sd(diff(spread),na.rm = T)
}


```


```{r}
#| label: calc-features

# I need a function that will calculate all features for each month end over the entire data period.
# the output should have half-life, hurst, and vol at month end for each pair
# it will be used for clustering and rebalancing at month end


month_ends <- dat_wide %>%
  dplyr::mutate(ym = format(date, "%Y-%m")) %>%
  dplyr::group_by(ym) %>%
  dplyr::slice_max(date, n = 1) %>%
  dplyr::ungroup() %>%
  dplyr::filter(date >= as.Date(train_start), date <= as.Date(test_end)) %>% 
  dplyr::pull(date)
  
compute_pair_feats_monthEnd <- function(pairs, month_end, wide_prices){
    leg1 <- pairs$leg1
    leg2 <- pairs$leg2
    valid_from <- pairs$valid_from

    window_roll_data <- wide_prices %>%
      dplyr::filter(date <= month_end & date >= valid_from) %>%
      dplyr::select(date, p1 = dplyr::all_of(leg1), p2 = dplyr::all_of(leg2)) %>%
      tidyr::drop_na() %>%
      dplyr::slice_tail(n = window_roll)

    if (nrow(window_roll_data) < 20) {
      return(tibble::tibble(
        pair_id = pairs$pair_id, date = month_end,
        half_life = NA_real_, hurst = NA_real_, spread_vol = NA_real_
      ))
    }

    beta_ols <- coef(lm(p1 ~ p2, data = window_roll_data))[2]

    spread <- window_roll_data$p1 - beta_ols * window_roll_data$p2

    tibble::tibble(
      pair_id    = pairs$pair_id,
      date       = month_end,
      half_life  = half_life(spread),
      hurst      = hurst_exp(spread),
      spread_vol = spread_vol(spread)
    )
  }

```

```{r}
#| label: month-end-feats

cl <- parallel::makeCluster(parallel::detectCores() - 1)
doParallel::registerDoParallel(cl)

task_grid <- tidyr::expand_grid(
  pair_pos = seq_len(nrow(pair_grid)),
  month_pos = seq_along(month_ends)
)

features_list <- foreach::foreach(
  i = seq_len(nrow(task_grid)),
  .combine = dplyr::bind_rows,
  .packages = c("tidyverse", "slider","pracma"),
  .export = c("pair_grid", "month_ends", "dat_wide", "window_roll",
              "half_life", "hurst_exp", "spread_vol",
              "compute_pair_feats_monthEnd")
) %dopar% {
  pr <- pair_grid[task_grid$pair_pos[i], ]
  md <- month_ends[task_grid$month_pos[i]]
  compute_pair_feats_monthEnd(pr, md, dat_wide)
}

parallel::stopCluster(cl)

# Clean: drop rows with any NA feature
features_clean <- features_list %>%
  tidyr::drop_na(half_life, hurst, spread_vol) %>%
  dplyr::filter(half_life > 0, half_life < Inf, spread_vol > 0)
```


### 3.2 Train/Test Split & Feature Normalization

```{r}
#| label: tidymodels-recipe

features_train <- features_clean %>%
  dplyr::filter(date <= train_end)

features_test <- features_clean %>%
  dplyr::filter(date >= test_start)


norm_recipe <- recipes::recipe(
  ~ half_life + hurst + spread_vol,
  data = features_train
) %>%
  recipes::step_normalize(recipes::all_numeric_predictors()) %>%
  recipes::prep()

train_baked <- recipes::bake(norm_recipe, features_train) %>%
  dplyr::bind_cols(features_train %>% dplyr::select(pair_id, date))

test_baked <- recipes::bake(norm_recipe, features_test) %>%
  dplyr::bind_cols(features_test %>% dplyr::select(pair_id, date))
```


### 3.3 Clustering: K Selection

```{r}
#| label: selecting-K

last_train_month <- max(features_train$date) # use last month of training - arbitrary

feat_snapshot <- train_baked %>%
  dplyr::filter(date == last_train_month) %>%
  dplyr::select(half_life, hurst, spread_vol)

feat_matrix <- as.matrix(feat_snapshot)

k_results <- tibble::tibble(k = k_min:k_max+1) %>%
  dplyr::mutate(
    km_fit = purrr::map(k, ~ kmeans(feat_matrix, centers = .x, nstart = 25, iter.max = 100)),
    
    wss = purrr::map_dbl(km_fit, ~ .x$tot.withinss),
    
    sil_km = purrr::map_dbl(k, function(kk) {
      
      km <- kmeans(feat_matrix, centers = kk, nstart = 25, iter.max = 100)
      if (nrow(feat_matrix) <= kk) return(NA_real_)
      mean(cluster::silhouette(km$cluster, dist(feat_matrix))[, 3])
    })
    ,
    sil_hc = purrr::map_dbl(k, function(kk) {
      hc <- hclust(dist(feat_matrix), method = "ward.D2")
      cl <- cutree(hc, k = kk)
      if (nrow(feat_matrix) <= kk) return(NA_real_) #had to add safeguard here
      mean(cluster::silhouette(cl, dist(feat_matrix))[, 3])
    })
  )

k_results <- k_results %>%
  dplyr::mutate(avg_sil = (sil_km + sil_hc) / 2)

optimal_k <- k_results %>% # max average k-means and h-clustering silhoutte
  dplyr::slice_max(avg_sil, n = 1) %>%
  dplyr::pull(k)

p_elbow <- ggplot(k_results, aes(x = k, y = wss)) + # elbow graph for presentation
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  labs(title = "Elbow Method", x = "k", y = "Total WSS") +
  theme_minimal()

p_sil <- k_results %>%
  tidyr::pivot_longer(cols = c(sil_km, sil_hc), names_to = "method",
                      values_to = "silhouette") %>%
  dplyr::mutate(method = ifelse(method == "sil_km", "K-Means", "Hierarchical")) %>%
  ggplot(aes(x = k, y = silhouette, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  labs(title = "Silhouette Score", x = "k", y = "Avg Silhouette Width") +
  theme_minimal()

p_elbow / p_sil


```


### 3.4 Tradeable Cluster Selection Function

+ The tradeable cluster is defined as the one whose centroid has the lowest combined rank of half-life and Hurst exponent. Low half-life means fast reversion; low Hurst (< 0.5) means mean-reverting structure. This function is called monthly during backtesting


```{r}
#| label: Cluster-selection

# cluster all etfs by half-life, hurst, and spread vol. using k=6
# trade within clusters that have lowest half-life and hurst (mean reverts quickly)

select_tradeable_pairs <- function(feat_df, k, top_n = top_pairs) {
  # feat_df must have columns: pair_id, half_life, hurst, spread_vol (raw)
  # plus normalized versions for clustering

  feat_mat <- feat_df %>%
    dplyr::select(half_life_norm, hurst_norm, spread_vol_norm) %>%
    as.matrix() 

  # K-means clustering

  km <- kmeans(feat_mat, centers = k, nstart = 25, iter.max = 100)
  feat_df$cluster <- km$cluster

  # Rank clusters by centroid
  centroids <- feat_df %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(
      mean_hl    = mean(half_life, na.rm = TRUE),
      mean_hurst = mean(hurst, na.rm = TRUE),
      mean_vol   = mean(spread_vol, na.rm = TRUE),
      n_pairs    = dplyr::n(),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      rank_hl    = rank(mean_hl), # simple but soo cool!
      rank_hurst = rank(mean_hurst),
      combined_rank = rank_hl + rank_hurst
    )

  best_cluster <- centroids %>%
    dplyr::slice_min(combined_rank, n = 1, with_ties = FALSE) %>%
    dplyr::pull(cluster)

  # Pairs in tradeable cluster
  tradeable <- feat_df %>%
    dplyr::filter(cluster == best_cluster) %>%
    dplyr::arrange(spread_vol) %>%
    dplyr::slice_head(n = min(top_n, nrow(.))) # if less than 10, we wont be under-invested

  # Inverse volatility parity weights - i will use this in trading to keep risk from being concentrated in one pair
  tradeable <- tradeable %>%
    dplyr::mutate(
      inv_vol = 1 / spread_vol,
      weight  = inv_vol / sum(inv_vol)
    )

  list(
    tradeable = tradeable,
    centroids = centroids,
    best_cluster = best_cluster,
    all_pairs = feat_df
  )
}


```



### 3.5 Cluster Visualization


```{r}
#| label: cluster-viz

# Visualize clusters for the last training month
snap_raw <- features_train %>%
  dplyr::filter(date == last_train_month)

snap_norm <- train_baked %>%
  dplyr::filter(date == last_train_month)

snap_combined <- snap_raw %>%
  dplyr::inner_join(
    snap_norm %>%
      dplyr::rename(half_life_norm = half_life, hurst_norm = hurst,
                     spread_vol_norm = spread_vol),
    by = c("pair_id", "date")
  )

cluster_result <- select_tradeable_pairs(snap_combined, optimal_k)

 
viz_df <- cluster_result$all_pairs %>%
    dplyr::mutate(
      is_tradeable = cluster == cluster_result$best_cluster,
      cluster_label = factor(cluster)
    )

  p_hl_hurst <- ggplot(viz_df, aes(x = half_life, y = hurst,
                                     color = cluster_label, shape = is_tradeable)) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = "Clusters: Half-Life vs Hurst", x = "Half-Life (days)",
         y = "Hurst Exponent", color = "Cluster", shape = "Tradeable") +
    theme_minimal()

  p_hl_vol <- ggplot(viz_df, aes(x = half_life, y = spread_vol,
                                   color = cluster_label, shape = is_tradeable)) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = "Clusters: Half-Life vs Spread Vol", x = "Half-Life (days)",
         y = "Spread Volatility", color = "Cluster", shape = "Tradeable") +
    theme_minimal()

  p_hl_hurst / p_hl_vol

```


```{r}
#| label: cluster-selection-table

cluster_result$centroids %>%
  dplyr::mutate(selected = ifelse(cluster == cluster_result$best_cluster,
                                  "*** Tradeable ***", "")) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(mean_hl, mean_hurst, mean_vol,
                                rank_hl, rank_hurst, combined_rank), decimals = 2) %>%
  gt::tab_header(title = "Cluster Centroids (Last Training Month)")

```


## 4. Layer 2: Signal Generation (Kalman Filter)

### 4.1 Kalman Filter Setup

+ For each active pair, we fit a Kalman Filter that tracks how the relationship between the two assets evolves over time. Instead of assuming a fixed hedge ratio like OLS does, the KF lets the ratio drift day by day. The filter also estimates a dynamic intercept to account for persistent price level differences. Each day it outputs an updated hedge ratio, a spread estimate, and a prediction error (innovation) that tells us how surprised the model was by today's prices.


```{r}
#| label: Kalman-filtering

# Build dlm model for a pair
build_dlm <- function(par, X) { 
  
# initalize the model for filtering. this allows dynamic regression (i.e incorporating new data as it comes in to the 
# new lm fit)
# this model only takes the independent variable, then we feed it the dependent later on
  
  dlm::dlmModReg(
    X  = matrix(X, ncol = 1),
    dV = exp(par[1]),
    dW = exp(par[2:3])
  )
}

# Fit Kalman Filter for a single pair over a price vector

fit_kalman_pair <- function(y, X, train_end_pos = NULL) {
  # y = price of leg1, X = price of leg2
  n <- length(y)
  if (is.null(train_end_pos)) train_end_pos <- n

  y_train <- y[1:train_end_pos]
  X_train <- X[1:train_end_pos]

  # small init state noise biases optimizer toward smoother beta paths
  init_parm <- c(log(var(y_train, na.rm = TRUE)), log(1e-4), log(1e-4))

  fit <- tryCatch(
    dlm::dlmMLE(
      y      = y_train,
      parm   = init_parm,
      build  = function(par) build_dlm(par, X_train),
      method = "L-BFGS-B",
      lower  = c(-Inf, -20, -20),
      upper  = c( Inf,  -4,  -4),  # cap state noise at exp(-4) ≈ 0.018 to prevent beta_t drift
      control = list(maxit = 500)
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) return(NULL)

  mod  <- build_dlm(fit$par, X) # using optimal paramters found above
  filt <- dlm::dlmFilter(y, mod) # this runs on all data - including testing, but using params from training

  alpha_t <- as.numeric(filt$m[-1, 1])  # intercept is column 1 in dlmModReg
  beta_t  <- as.numeric(filt$m[-1, 2])  # hedge ratio is column 2
  spread  <- y - beta_t * X - alpha_t # this is the dynamic spread built from the output of filtering

  # Innovations (one-step prediction errors)
  y_pred <- as.numeric(filt$f)
  innov  <- y - y_pred # difference between actual and expected

  tibble::tibble(
    pos     = seq_len(n),
    beta_t  = beta_t,
    alpha_t = alpha_t,
    spread  = spread,
    innov   = innov
  )
}


```




### 4.2 Signal Generation

Trade signals fire when two conditions are met simultaneously: (1) the pair is in the current month's tradeable cluster, and (2) the Kalman Filter spread deviates beyond `entry_zscore` standard deviations of its rolling history. Long the pair when the spread is negative (leg1 undervalued relative to leg2), short when spread is positive.


```{r}
#| label: signal-generation

generate_signals <- function(kf_result, entry_thresh = entry_zscore,
                             roll_sd_window = 60L) {
  # kf_result is a tibble with columns: idx, beta_t, alpha_t, spread, innov

  kf_result %>%
    dplyr::mutate(
      spread_roll_mean = slider::slide_dbl(
        spread, mean, .before = roll_sd_window - 1, .complete = TRUE
      ),
      spread_roll_sd = slider::slide_dbl(
        spread, sd, .before = roll_sd_window - 1, .complete = TRUE
      ),
      z_score = (spread - spread_roll_mean) / spread_roll_sd,
      # Innovation spike detection
      innov_roll_mean = slider::slide_dbl(
        abs(innov), mean, .before = roll_sd_window - 1, .complete = TRUE
      ),
      innov_roll_sd = slider::slide_dbl(
        abs(innov), sd, .before = roll_sd_window - 1, .complete = TRUE
      ),
      innov_spike = abs(innov) > (innov_roll_mean + innov_mult * innov_roll_sd)
    )
}


```



## 5. Exit Rules & Override

### 5.1 Exit Logic

+ Three exit conditions govern position closure. Take-profit triggers when the spread returns to the Kalman Filter mean estimate (z-score crosses zero). The time stop exits after `2 * half_life` days regardless of P&L, preventing capital from being trapped in broken pairs. The innovation override force-exits when the KF prediction errors spike for `max_innov_days` consecutive days, signaling a structural break in the pair relationship.


## 6. Monthly Rebalancing & Full Backtest

### 6.1 Backtesting

+ The backtest iterates month by month. At each month-end: Layer 1 re-clusters all 66 pairs and selects the tradeable set with inverse volatility parity weights. During the month, Layer 2 runs the Kalman Filter daily on each active pair. Pairs leaving the tradeable cluster are force-closed; pairs remaining are resized to new weights.


```{r}
#| label: backtest-function
#| cache: true
run_backtest <- function(pw, pair_grid, feat_data_raw, feat_data_norm,
                           norm_rec, k, entry_thresh = entry_zscore,
                           dd_threshold = NULL,
                           start_date, end_date) {

    cl <- parallel::makeCluster(parallel::detectCores() - 1)
    doParallel::registerDoParallel(cl)
    on.exit(parallel::stopCluster(cl))

    trading_dates <- pw %>%
      dplyr::filter(date >= start_date, date <= end_date) %>%
      dplyr::pull(date) %>%
      sort()

    month_ends_bt <- pw %>%
      dplyr::filter(date >= start_date - 60, date <= end_date) %>%
      dplyr::mutate(ym = format(date, "%Y-%m")) %>%
      dplyr::group_by(ym) %>%
      dplyr::slice_max(date, n = 1) %>%
      dplyr::ungroup() %>%
      dplyr::pull(date) %>%
      sort()

    pre_start_months <- month_ends_bt[month_ends_bt < start_date]
    in_range_months  <- month_ends_bt[month_ends_bt >= start_date & month_ends_bt <=
  end_date]
    rebal_dates      <- c(tail(pre_start_months, 1), in_range_months)

    daily_results <- tibble::tibble(
      date     = trading_dates,
      port_ret = 0
    )

    current_pairs <- NULL

    for (rd in seq_along(rebal_dates)) {
      rebal_date <- rebal_dates[rd]
      next_rebal <- if (rd < length(rebal_dates)) rebal_dates[rd + 1] else end_date

      feat_raw_month  <- feat_data_raw  %>% dplyr::filter(date == rebal_date)
      feat_norm_month <- feat_data_norm %>% dplyr::filter(date == rebal_date)

      if (nrow(feat_raw_month) < k || nrow(feat_norm_month) < k) next

      feat_combined <- feat_raw_month %>%
        dplyr::inner_join(
          feat_norm_month %>%
            dplyr::rename(half_life_norm = half_life, hurst_norm = hurst,
                          spread_vol_norm = spread_vol),
          by = c("pair_id", "date")
        )

      cl_result <- select_tradeable_pairs(feat_combined, k, top_pairs)
      if (is.null(cl_result)) next

      current_pairs <- cl_result$tradeable

      month_dates <- trading_dates[trading_dates > rebal_date &
                                   trading_dates <= next_rebal]
      if (length(month_dates) == 0) next

      # Portfolio drawdown circuit breaker: pause new entries if drawdown exceeds threshold
      if (!is.null(dd_threshold) && any(daily_results$port_ret != 0)) {
        cum_rets   <- cumprod(1 + daily_results$port_ret)
        current_dd <- (tail(cum_rets, 1) - max(cum_rets)) / max(cum_rets)
        if (current_dd < dd_threshold) next
      }

      kf_lookback <- 252L
      kf_start    <- max(1L, which(pw$date <= rebal_date) %>% max() - kf_lookback)

      # Parallel pair loop — each worker returns month_rets or NULL
      pair_rets_list <- foreach::foreach(
        i = seq_len(nrow(current_pairs)),
        .packages = c("dplyr", "tidyr", "dlm", "slider", "tibble"),
        .export  = c("fit_kalman_pair", "generate_signals", "build_dlm",
                     "innov_mult", "max_innov_days"),
        .errorhandling = "remove"
      ) %dopar% {

        pair_info <- current_pairs[i, ]
        legs <- strsplit(pair_info$pair_id, "_")[[1]]
        leg1 <- legs[1]; leg2 <- legs[2]

        pair_prices <- pw %>%
          dplyr::slice(kf_start:nrow(pw)) %>%
          dplyr::filter(date <= next_rebal) %>%
          dplyr::select(date, p1 = dplyr::all_of(leg1), p2 = dplyr::all_of(leg2)) %>%
          tidyr::drop_na() %>%
          dplyr::mutate(
            p1_ret = p1 / dplyr::lag(p1) - 1,
            p2_ret = p2 / dplyr::lag(p2) - 1
          )

        if (nrow(pair_prices) < 252) return(NULL)

        training_end <- max(which(pair_prices$date <= rebal_date))

        kf_out <- fit_kalman_pair(
          y             = pair_prices$p1,
          X             = pair_prices$p2,
          train_end_pos = training_end
        )

        if (is.null(kf_out)) return(NULL)

        kf_out$date <- pair_prices$date

        kf_signals <- generate_signals(kf_out, entry_thresh = entry_thresh)

        half_life_val <- pair_info$half_life
        if (is.na(half_life_val) || half_life_val <= 0) half_life_val <- 20

        time_stop_days <- max(round(2 * half_life_val), 5)

        kf_signals <- kf_signals %>%
          dplyr::mutate(
            raw_signal = dplyr::case_when(
              is.na(z_score)          ~ 0L,
              z_score < -entry_thresh ~ 1L,
              z_score >  entry_thresh ~ -1L,
              TRUE                    ~ 0L
            ),
            innov_consec = slider::slide_dbl(
              as.numeric(dplyr::coalesce(innov_spike, FALSE)),
              sum, .before = max_innov_days - 1, .complete = TRUE
            ),
            innov_override = dplyr::coalesce(innov_consec >= max_innov_days, FALSE)
          )

        n_sig         <- nrow(kf_signals)
        position      <- rep(0L, n_sig)
        days_in_trade <- rep(0L, n_sig)

        for (t in 2:n_sig) {
          prev_pos <- position[t - 1]
          sig      <- kf_signals$raw_signal[t]

          if (prev_pos != 0) {
            days_in_trade[t] <- days_in_trade[t - 1] + 1L

            z_now       <- kf_signals$z_score[t]
            take_profit <- !is.na(z_now) && (
              (prev_pos > 0 && z_now >= 0) || (prev_pos < 0 && z_now <= 0)
            )
            time_stop  <- days_in_trade[t] > time_stop_days
            innov_exit <- isTRUE(kf_signals$innov_override[t])

            if (take_profit || time_stop || innov_exit) {
              position[t]      <- 0L
              days_in_trade[t] <- 0L
              next
            }

            position[t] <- prev_pos
          } else {
            if (!is.na(sig) && sig != 0) {
              position[t]      <- sig
              days_in_trade[t] <- 1L
            }
          }
        }

        kf_signals$position <- position

        kf_signals <- kf_signals %>%
          dplyr::left_join(
            pair_prices %>% dplyr::select(date, p1_ret, p2_ret),
            by = "date"
          ) %>%
          dplyr::mutate(
            pair_ret = dplyr::lag(position) * (p1_ret - dplyr::lag(beta_t) * p2_ret),
          pair_ret = pmax(pmin(pair_ret, 0.25), -0.25)
          )

        kf_signals %>%
          dplyr::filter(date %in% month_dates) %>%
          dplyr::select(date, pair_ret) %>%
          dplyr::mutate(pair_ret = dplyr::coalesce(pair_ret, 0) * pair_info$weight)
      }

      # Aggregate all pair contributions for this month into daily_results
      valid_rets <- Filter(function(x) is.data.frame(x) && nrow(x) > 0, pair_rets_list)
      if (length(valid_rets) == 0) next

      month_contributions <- dplyr::bind_rows(valid_rets) %>%
        dplyr::group_by(date) %>%
        dplyr::summarise(port_ret_add = sum(pair_ret, na.rm = TRUE), .groups = "drop")

      daily_results <- daily_results %>%
        dplyr::left_join(month_contributions, by = "date") %>%
        dplyr::mutate(
          port_ret = port_ret + dplyr::coalesce(port_ret_add, 0)
        ) %>%
        dplyr::select(-port_ret_add)
    }

    daily_results %>%
      dplyr::mutate(cumeq = cumprod(1 + port_ret))
  }

  # Run backtest on training period
  train_results <- run_backtest(
    pw = dat_wide,
    pair_grid = pair_grid,
    feat_data_raw = features_clean,
    feat_data_norm = train_baked %>%
      dplyr::rename(half_life = half_life, hurst = hurst, spread_vol = spread_vol),
    norm_rec = norm_recipe,
    k = optimal_k,
    entry_thresh = entry_zscore,
    start_date = train_start,
    end_date = train_end
  )

```


## 7. Performance Analysis

### 7.1 Backtest Runs

```{r}
#| label: backtest-test
#| cache: true

test_results <- run_backtest(
  pw             = dat_wide,
  pair_grid      = pair_grid,
  feat_data_raw  = features_clean,
  feat_data_norm = test_baked,
  norm_rec       = norm_recipe,
  k              = optimal_k,
  entry_thresh   = entry_zscore,
  start_date     = test_start,
  end_date       = test_end
)
```

### 7.2 Performance Metrics

```{r}
#| label: performance-metrics

spy_prices_full <- dat_raw %>%
  dplyr::filter(ticker == "SPY") %>%
  dplyr::select(date, price) %>%
  dplyr::mutate(spy_ret = price / dplyr::lag(price) - 1)

tr_xts <- train_results %>% dplyr::select(date, port_ret) %>% timetk::tk_xts(date_var = date)
te_xts <- test_results  %>% dplyr::select(date, port_ret) %>% timetk::tk_xts(date_var = date)

spy_tr_xts <- spy_prices_full %>%
  dplyr::filter(date >= train_start, date <= train_end) %>%
  timetk::tk_xts(date_var = date)

spy_te_xts <- spy_prices_full %>%
  dplyr::filter(date >= test_start, date <= test_end) %>%
  timetk::tk_xts(date_var = date)

tibble::tibble(
  Period         = c("Training", "Test"),
  `Cum. Return`  = c(tail(train_results$cumeq,1)-1, tail(test_results$cumeq,1)-1),
  `Sharpe`       = c(SharpeRatio(tr_xts, annualize=TRUE)[1], SharpeRatio(te_xts, annualize=TRUE)[1]),
  `Omega Sharpe` = c(OmegaSharpeRatio(as.ts(tr_xts), MAR=0), OmegaSharpeRatio(as.ts(te_xts), MAR=0)),
  `Max Drawdown` = c(maxDrawdown(tr_xts), maxDrawdown(te_xts)),
  `SPY Corr`     = c(cor(tr_xts, spy_tr_xts, use="complete.obs")[1,2],
                     cor(te_xts, spy_te_xts, use="complete.obs")[1,2])
) %>%
  gt::gt() %>%
  gt::tab_header(title = "Strategy Performance vs SPY") %>%
  gt::fmt_percent(columns = c(`Cum. Return`, `Max Drawdown`), decimals = 1) %>%
  gt::fmt_number(columns = c(Sharpe, `Omega Sharpe`, `SPY Corr`), decimals = 3) %>%
  gt::tab_style(
    style     = gt::cell_text(weight = "bold"),
    locations = gt::cells_column_labels()
  )
```

### 7.3 Equity Curves



```{r}
#| label: equity-curves-test

spy_test <- spy_prices_full %>%
  dplyr::filter(date >= test_start, date <= test_end) %>%
  dplyr::mutate(cumeq = cumprod(1 + tidyr::replace_na(spy_ret, 0)))

dplyr::bind_rows(
  test_results %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "Strategy"),
  spy_test     %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "SPY")
) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq, color = Series)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_color_manual(values = c("Strategy" = "black", "SPY" = "firebrick")) +
  ggplot2::labs(title = "Test Period: Strategy vs SPY", x = NULL, y = "Cumulative Equity", color = NULL) +
  ggplot2::theme_minimal()
```

```{r}
#| label: equity-curves-train

spy_train <- spy_prices_full %>%
  dplyr::filter(date >= train_start, date <= train_end) %>%
  dplyr::mutate(cumeq = cumprod(1 + tidyr::replace_na(spy_ret, 0)))

dplyr::bind_rows(
  train_results %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "Strategy"),
  spy_train     %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "SPY")
) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq, color = Series)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_color_manual(values = c("Strategy" = "black", "SPY" = "firebrick")) +
  ggplot2::labs(title = "Training Period: Strategy vs SPY", x = NULL, y = "Cumulative Equity", color = NULL) +
  ggplot2::theme_minimal()
```

### 7.4 Sensitivity: Excluding the 2020 COVID Shock

+ The COVID crash (February–June 2020) was an extreme systemic event where all sector ETFs fell simultaneously, violating the mean-reversion assumption across all pairs at once. The following shows what the training period equity curve would look like had the strategy been paused during this window.

```{r}
#| label: no2020

train_results_no2020 <- train_results %>%
  dplyr::mutate(
    port_ret = ifelse(date >= as.Date("2020-02-20") & date <= as.Date("2020-06-30"), 0, port_ret),
    cumeq    = cumprod(1 + port_ret)
  )

dplyr::bind_rows(
  train_results        %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "Strategy"),
  train_results_no2020 %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "Strategy (ex-2020)"),
  spy_train            %>% dplyr::select(date, cumeq) %>% dplyr::mutate(Series = "SPY")
) %>%
  ggplot2::ggplot(ggplot2::aes(x = date, y = cumeq, color = Series)) +
  ggplot2::geom_line(linewidth = 0.8) +
  ggplot2::scale_color_manual(
    values = c("Strategy" = "black", "Strategy (ex-2020)" = "steelblue", "SPY" = "firebrick")
  ) +
  ggplot2::labs(title = "Training Period: Impact of 2020 COVID Shock",
                x = NULL, y = "Cumulative Equity", color = NULL) +
  ggplot2::theme_minimal()
```

## 8. Learnings

+ **Market neutrality holds in training but breaks in the test period.** The strategy achieved near-zero SPY correlation (-0.030) during training, confirming the pairs construction produces genuinely market-neutral exposures. However, correlation rose to 0.364 in the test period, driven by the April 2025 tariff shock which caused all sector ETFs to move together — collapsing the cross-sectional dispersion that pairs trading requires.

+ **The strategy fails to generate positive alpha.** Negative Sharpe and Omega Sharpe ratios in both periods indicate the strategy loses money on a risk-adjusted basis. The training period cumulative loss of -31.8% stems primarily from a sustained post-2010 decline as sector ETF correlations tightened in the low-volatility regime. The strategy works best in high-dispersion environments (2008) and struggles in trending, correlated markets.

+ **Systemic shocks overwhelm pair-level exit rules.** The innovation override and time stop are designed for gradual structural breaks within individual pairs. During COVID (March 2020) and the 2025 tariff shock, the breakdown was simultaneous across all pairs and faster than daily-frequency rules can respond. A portfolio-level drawdown circuit breaker was explored but did not improve results materially — the losses were too persistent for a threshold-based halt to be effective.

+ **KF beta instability was a significant implementation challenge.** The DLM state vector column ordering (intercept first, slope second) was initially inverted, causing the hedge ratio to be estimated at 30–370x its true value and producing extreme position sizes. Constraining the MLE optimizer (L-BFGS-B with upper bounds on state noise) and correcting the column assignment resolved this. Future implementations should validate beta estimates against OLS as a sanity check before trading.

+ **Return computation uses adjusted close-to-close prices throughout.** A more precise implementation would distinguish entry days (open price) from hold days (close-to-close), consistent with the actual execution assumption. At daily frequency the impact is small but non-zero, and represents a known limitation of the current backtest.