---
title: "Clustering-Based Pairs Trading on S&P 500 Sector ETFs"
subtitle: "FIN 452 — Quantitative Trading Project"
author: "Soumer"
date: "`r Sys.Date()`" 
format:
  html:
    embed-resources: true
    toc: true
    toc-depth: 3
    code-fold: true
    code-tools: true
    theme: flatly
    fig-width: 10
    fig-height: 6
execute:
  warning: false
  message: false
  echo: false
---






## 1. Executive Summary

 ![Strategy Mental Model](assets/mental_model.svg)
 
 
+ This project implements a two-layer clustering-based pairs trading strategy on 12 S&P 500 Select Sector SPDR ETFs. Layer 1 selects tradeable pairs monthly via k-means and hierarchical clustering on mean-reversion features (half-life, Hurst exponent, spread volatility). Layer 2 generates daily trade signals using a Kalman Filter that tracks dynamic hedge ratios and spread deviations. The strategy targets risk-adjusted outperformance of the SPY benchmark over the January 2025 -- March 2026 test period.

## 2. Universe & Data

```{r}
#| label: Libraries

library(tidyverse)
library(RTL)
library(xts)
library(TTR)
library(tidyquant)
library(pracma)
library(doParallel)
library(foreach)
library(parallel)
library(gt)
library(patchwork)
library(dlm)
```




```{r}
#| label: Parameters

train_start  <- as.Date("2000-06-01")
train_end    <- as.Date("2024-12-31")
test_start   <- as.Date("2025-01-01")
test_end    <- as.Date("2026-03-31")
window_roll  <- 252L
entry_zscore <- 1.5
max_innov_days <- 5L
innov_mult   <- 3.0
top_pairs  <- 10L
k_min        <- 2L
k_max        <- 12L
XLC_inception <- as.Date("2018-06-19")
XLRE_inception <- as.Date("2015-10-07")
```
### 2.1 Assets Considered

+ The universe consists of 11 sector ETFs plus SPY as both benchmark and tradeable asset. XLC (Communication Services) is only available from June 2018 and XLRE (Real Estate) from October 2015. Pairs involving these tickers are excluded before their inception dates.

```{r}
#| label: etfs-info-table
#| 
tickers <- c("SPY","XLB","XLC","XLE","XLF","XLI","XLK","XLP","XLRE","XLV","XLU","XLY")

ticker_info <- tibble::tibble(
  ticker = tickers,
  sector = c("S&P 500 Benchmark", "Materials", "Communication Services",
             "Energy", "Financials", "Industrials", "Technology",
             "Consumer Staples", "Real Estate", "Health Care",
             "Utilities", "Consumer Discretionary"),
  inception_note = c("", "", "From 2018-06-19", "", "", "", "",
                     "", "From 2015-10-07", "", "", "")
)

ticker_info %>% gt::gt() %>%
  gt::tab_header(title = "Asset Universe") %>%
  gt::cols_label(ticker = "Ticker", sector = "Sector",
                 inception_note = "Note")
```


```{r}
#| label: Data Download

dat_raw <- tidyquant::tq_get(tickers,
                             get = "stock.prices",
                             from = train_start,
                             to = test_end + 1) %>% 
  dplyr::select(date,symbol,adjusted) %>% 
  dplyr::rename(ticker = symbol, price = adjusted)

dat_wide <- dat_raw %>% 
  tidyr::pivot_wider(names_from = ticker,
                     values_from = price) %>% 
  dplyr::arrange(date)

spy_prices <- dat_raw %>%
  dplyr::filter(ticker == "SPY") %>%
  dplyr::select(date, price) %>%
  dplyr::mutate(spy_ret = price / dplyr::lag(price) - 1)


```

### 2.2 Pair Definitions

+ We enumerate all 66 unique pairs from the 12 assets. Each pair has a validity start date determined by the later inception date of its two legs.

```{r}
#| label: Pair Construction

pair_grid <- tidyr::expand_grid(
  leg1 = tickers,
  leg2 = tickers
) %>%
  dplyr::filter(leg1 < leg2) %>%
  dplyr::mutate(
    pair_id = paste0(leg1, "_", leg2),
    valid_from = dplyr::case_when(
      leg1 == "XLC" | leg2 == "XLC" ~ XLC_inception,
      leg1 == "XLRE" | leg2 == "XLRE" ~ XLRE_inception,
      TRUE ~ train_start
    )
  )
```


## 3. Layer 1: Monthly Pair Selection

### 3.1 Spread Feature Computation

+ For each of the 66 pairs, we compute three rolling features that characterize mean-reversion behavior. The spread is defined as `S_t = p1_t - beta_hat * p2_t` where `beta_hat` comes from an OLS regression over the rolling window_roll. Half-life measures speed of reversion, Hurst exponent measures persistence structure (H < 0.5 is mean-reverting), and spread volatility captures the magnitude of movements.

```{r}
#| label: Features-funcs

half_life <- function(spread) {
  
  n <- length(spread)
  ds <- diff(spread)
  s_lag <- spread[-n]
  
  fit <- lm(ds ~ s_lag)
  beta <- coef(fit)[2]
  -log(2) / beta
}


hurst_exp <- function(spread){
  
  pracma::hurstexp(spread,d = 20,display = F)$Hrs
}


spread_vol <- function(spread){
  
  sd(diff(spread),na.rm = T)
}


```


```{r}
#| label: calc-features

# I need a function that will calculate all features for each month end over the entire data period.
# the output should have half-life, hurst, and vol at month end for each pair
# it will be used for clustering and rebalancing at month end


month_ends <- dat_wide %>%
  dplyr::mutate(ym = format(date, "%Y-%m")) %>%
  dplyr::group_by(ym) %>%
  dplyr::slice_max(date, n = 1) %>%
  dplyr::ungroup() %>%
  dplyr::filter(date >= as.Date(train_start), date <= as.Date(test_end)) %>% 
  dplyr::pull(date)
  
compute_pair_feats_monthEnd <- function(pairs, month_end, wide_prices){
    leg1 <- pairs$leg1
    leg2 <- pairs$leg2
    valid_from <- pairs$valid_from

    window_roll_data <- wide_prices %>%
      dplyr::filter(date <= month_end & date >= valid_from) %>%
      dplyr::select(date, p1 = dplyr::all_of(leg1), p2 = dplyr::all_of(leg2)) %>%
      tidyr::drop_na() %>%
      dplyr::slice_tail(n = window_roll)

    if (nrow(window_roll_data) < 20) {
      return(tibble::tibble(
        pair_id = pairs$pair_id, date = month_end,
        half_life = NA_real_, hurst = NA_real_, spread_vol = NA_real_
      ))
    }

    beta_ols <- coef(lm(p1 ~ p2, data = window_roll_data))[2]

    spread <- window_roll_data$p1 - beta_ols * window_roll_data$p2

    tibble::tibble(
      pair_id    = pairs$pair_id,
      date       = month_end,
      half_life  = half_life(spread),
      hurst      = hurst_exp(spread),
      spread_vol = spread_vol(spread)
    )
  }

```

```{r}
#| label: month-end-feats

cl <- parallel::makeCluster(parallel::detectCores() - 1)
doParallel::registerDoParallel(cl)

task_grid <- tidyr::expand_grid(
  pair_pos = seq_len(nrow(pair_grid)),
  month_pos = seq_along(month_ends)
)

features_list <- foreach::foreach(
  i = seq_len(nrow(task_grid)),
  .combine = dplyr::bind_rows,
  .packages = c("tidyverse", "slider","pracma"),
  .export = c("pair_grid", "month_ends", "dat_wide", "window_roll",
              "half_life", "hurst_exp", "spread_vol",
              "compute_pair_feats_monthEnd")
) %dopar% {
  pr <- pair_grid[task_grid$pair_pos[i], ]
  md <- month_ends[task_grid$month_pos[i]]
  compute_pair_feats_monthEnd(pr, md, dat_wide)
}

parallel::stopCluster(cl)

# Clean: drop rows with any NA feature
features_clean <- features_list %>%
  tidyr::drop_na(half_life, hurst, spread_vol) %>%
  dplyr::filter(half_life > 0, half_life < Inf, spread_vol > 0)
```


### 3.2 Train/Test Split & Feature Normalization

```{r}
#| label: tidymodels-recipe

features_train <- features_clean %>%
  dplyr::filter(date <= train_end)

features_test <- features_clean %>%
  dplyr::filter(date >= test_start)


norm_recipe <- recipes::recipe(
  ~ half_life + hurst + spread_vol,
  data = features_train
) %>%
  recipes::step_normalize(recipes::all_numeric_predictors()) %>%
  recipes::prep()

train_baked <- recipes::bake(norm_recipe, features_train) %>%
  dplyr::bind_cols(features_train %>% dplyr::select(pair_id, date))

test_baked <- recipes::bake(norm_recipe, features_test) %>%
  dplyr::bind_cols(features_test %>% dplyr::select(pair_id, date))
```


### 3.3 Clustering: K Selection

```{r}
#| label: selecting-K

last_train_month <- max(features_train$date) # use last month of training - arbitrary

feat_snapshot <- train_baked %>%
  dplyr::filter(date == last_train_month) %>%
  dplyr::select(half_life, hurst, spread_vol)

feat_matrix <- as.matrix(feat_snapshot)

k_results <- tibble::tibble(k = k_min:k_max+1) %>%
  dplyr::mutate(
    km_fit = purrr::map(k, ~ kmeans(feat_matrix, centers = .x, nstart = 25, iter.max = 100)),
    
    wss = purrr::map_dbl(km_fit, ~ .x$tot.withinss),
    
    sil_km = purrr::map_dbl(k, function(kk) {
      
      km <- kmeans(feat_matrix, centers = kk, nstart = 25, iter.max = 100)
      if (nrow(feat_matrix) <= kk) return(NA_real_)
      mean(cluster::silhouette(km$cluster, dist(feat_matrix))[, 3])
    })
    ,
    sil_hc = purrr::map_dbl(k, function(kk) {
      hc <- hclust(dist(feat_matrix), method = "ward.D2")
      cl <- cutree(hc, k = kk)
      if (nrow(feat_matrix) <= kk) return(NA_real_) #had to add safeguard here
      mean(cluster::silhouette(cl, dist(feat_matrix))[, 3])
    })
  )

k_results <- k_results %>%
  dplyr::mutate(avg_sil = (sil_km + sil_hc) / 2)

optimal_k <- k_results %>% # max average k-means and h-clustering silhoutte
  dplyr::slice_max(avg_sil, n = 1) %>%
  dplyr::pull(k)

p_elbow <- ggplot(k_results, aes(x = k, y = wss)) + # elbow graph for presentation
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  labs(title = "Elbow Method", x = "k", y = "Total WSS") +
  theme_minimal()

p_sil <- k_results %>%
  tidyr::pivot_longer(cols = c(sil_km, sil_hc), names_to = "method",
                      values_to = "silhouette") %>%
  dplyr::mutate(method = ifelse(method == "sil_km", "K-Means", "Hierarchical")) %>%
  ggplot(aes(x = k, y = silhouette, color = method)) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_vline(xintercept = optimal_k, linetype = "dashed", color = "red") +
  labs(title = "Silhouette Score", x = "k", y = "Avg Silhouette Width") +
  theme_minimal()

p_elbow / p_sil


```


### 3.4 Tradeable Cluster Selection Function

+ The tradeable cluster is defined as the one whose centroid has the lowest combined rank of half-life and Hurst exponent. Low half-life means fast reversion; low Hurst (< 0.5) means mean-reverting structure. This function is called monthly during backtesting


```{r}
#| label: Cluster-selection

# cluster all etfs by half-life, hurst, and spread vol. using k=6
# trade within clusters that have lowest half-life and hurst (mean reverts quickly)

select_tradeable_pairs <- function(feat_df, k, top_n = top_pairs) {
  # feat_df must have columns: pair_id, half_life, hurst, spread_vol (raw)
  # plus normalized versions for clustering

  feat_mat <- feat_df %>%
    dplyr::select(half_life_norm, hurst_norm, spread_vol_norm) %>%
    as.matrix() 

  # K-means clustering

  km <- kmeans(feat_mat, centers = k, nstart = 25, iter.max = 100)
  feat_df$cluster <- km$cluster

  # Rank clusters by centroid
  centroids <- feat_df %>%
    dplyr::group_by(cluster) %>%
    dplyr::summarise(
      mean_hl    = mean(half_life, na.rm = TRUE),
      mean_hurst = mean(hurst, na.rm = TRUE),
      mean_vol   = mean(spread_vol, na.rm = TRUE),
      n_pairs    = dplyr::n(),
      .groups = "drop"
    ) %>%
    dplyr::mutate(
      rank_hl    = rank(mean_hl), # simple but soo cool!
      rank_hurst = rank(mean_hurst),
      combined_rank = rank_hl + rank_hurst
    )

  best_cluster <- centroids %>%
    dplyr::slice_min(combined_rank, n = 1, with_ties = FALSE) %>%
    dplyr::pull(cluster)

  # Pairs in tradeable cluster
  tradeable <- feat_df %>%
    dplyr::filter(cluster == best_cluster) %>%
    dplyr::arrange(spread_vol) %>%
    dplyr::slice_head(n = min(top_n, nrow(.))) # if less than 10, we wont be under-invested

  # Inverse volatility parity weights - i will use this in trading to keep risk from being concentrated in one pair
  tradeable <- tradeable %>%
    dplyr::mutate(
      inv_vol = 1 / spread_vol,
      weight  = inv_vol / sum(inv_vol)
    )

  list(
    tradeable = tradeable,
    centroids = centroids,
    best_cluster = best_cluster,
    all_pairs = feat_df
  )
}


```



### 3.5 Cluster Visualization


```{r}
#| label: cluster-viz

# Visualize clusters for the last training month
snap_raw <- features_train %>%
  dplyr::filter(date == last_train_month)

snap_norm <- train_baked %>%
  dplyr::filter(date == last_train_month)

snap_combined <- snap_raw %>%
  dplyr::inner_join(
    snap_norm %>%
      dplyr::rename(half_life_norm = half_life, hurst_norm = hurst,
                     spread_vol_norm = spread_vol),
    by = c("pair_id", "date")
  )

cluster_result <- select_tradeable_pairs(snap_combined, optimal_k)

 
viz_df <- cluster_result$all_pairs %>%
    dplyr::mutate(
      is_tradeable = cluster == cluster_result$best_cluster,
      cluster_label = factor(cluster)
    )

  p_hl_hurst <- ggplot(viz_df, aes(x = half_life, y = hurst,
                                     color = cluster_label, shape = is_tradeable)) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = "Clusters: Half-Life vs Hurst", x = "Half-Life (days)",
         y = "Hurst Exponent", color = "Cluster", shape = "Tradeable") +
    theme_minimal()

  p_hl_vol <- ggplot(viz_df, aes(x = half_life, y = spread_vol,
                                   color = cluster_label, shape = is_tradeable)) +
    geom_point(size = 3, alpha = 0.7) +
    labs(title = "Clusters: Half-Life vs Spread Vol", x = "Half-Life (days)",
         y = "Spread Volatility", color = "Cluster", shape = "Tradeable") +
    theme_minimal()

  p_hl_hurst / p_hl_vol

```


```{r}
#| label: cluster-selection-table

cluster_result$centroids %>%
  dplyr::mutate(selected = ifelse(cluster == cluster_result$best_cluster,
                                  "*** Tradeable ***", "")) %>%
  gt::gt() %>%
  gt::fmt_number(columns = c(mean_hl, mean_hurst, mean_vol,
                                rank_hl, rank_hurst, combined_rank), decimals = 2) %>%
  gt::tab_header(title = "Cluster Centroids (Last Training Month)")

```


## 4. Layer 2: Signal Generation (Kalman Filter)

### 4.1 Kalman Filter Setup

+ For each active pair, we fit a Kalman Filter that tracks how the relationship between the two assets evolves over time. Instead of assuming a fixed hedge ratio like OLS does, the KF lets the ratio drift day by day. The filter also estimates a dynamic intercept to account for persistent price level differences. Each day it outputs an updated hedge ratio, a spread estimate, and a prediction error (innovation) that tells us how surprised the model was by today's prices.


```{r}
#| label: Kalman-filtering

# Build dlm model for a pair
build_dlm <- function(par, X) { 
  
# initalize the model for filtering. this allows dynamic regression (i.e incorporating new data as it comes in to the 
# new lm fit)
# this model only takes the independent variable, then we feed it the dependent later on
  
  dlm::dlmModReg(
    X  = matrix(X, ncol = 1),
    dV = exp(par[1]),
    dW = exp(par[2:3])
  )
}

# Fit Kalman Filter for a single pair over a price vector

fit_kalman_pair <- function(y, X, train_end_pos = NULL) {
  # y = price of leg1, X = price of leg2
  n <- length(y)
  if (is.null(train_end_pos)) train_end_pos <- n

  y_train <- y[1:train_end_pos]
  X_train <- X[1:train_end_pos]

  init_parm <- c(log(var(y_train, na.rm = TRUE)), log(0.01), log(0.01)) 
  # giving it starting values to find dv and dW (basically works the same way as solver - using MLE)

  fit <- tryCatch(
    dlm::dlmMLE(
      y     = y_train,
      parm  = init_parm,
      build = function(par) build_dlm(par, X_train),
      control = list(maxit = 500) # limit so it doesnt run forever if it cant converge
    ),
    error = function(e) NULL
  )

  if (is.null(fit)) return(NULL)

  mod  <- build_dlm(fit$par, X) # using optimal paramters found above
  filt <- dlm::dlmFilter(y, mod) # this runs on all data - including testing, but using params from training

  beta_t  <- as.numeric(filt$m[-1, 1])
  alpha_t <- as.numeric(filt$m[-1, 2])
  spread  <- y - beta_t * X - alpha_t # this is the dynamic spread built from the output of filtering

  # Innovations (one-step prediction errors)
  y_pred <- as.numeric(filt$f)
  innov  <- y - y_pred # difference between actual and expected

  tibble::tibble(
    pos     = seq_len(n),
    beta_t  = beta_t,
    alpha_t = alpha_t,
    spread  = spread,
    innov   = innov
  )
}


```




### 4.2 Signal Generation

Trade signals fire when two conditions are met simultaneously: (1) the pair is in the current month's tradeable cluster, and (2) the Kalman Filter spread deviates beyond `entry_zscore` standard deviations of its rolling history. Long the pair when the spread is negative (leg1 undervalued relative to leg2), short when spread is positive.


```{r}
#| label: signal-generation

generate_signals <- function(kf_result, entry_thresh = entry_zscore,
                             roll_sd_window = 60L) {
  # kf_result is a tibble with columns: idx, beta_t, alpha_t, spread, innov

  kf_result %>%
    dplyr::mutate(
      spread_roll_mean = slider::slide_dbl(
        spread, mean, .before = roll_sd_window - 1, .complete = TRUE
      ),
      spread_roll_sd = slider::slide_dbl(
        spread, sd, .before = roll_sd_window - 1, .complete = TRUE
      ),
      z_score = (spread - spread_roll_mean) / spread_roll_sd,
      # Innovation spike detection
      innov_roll_mean = slider::slide_dbl(
        abs(innov), mean, .before = roll_sd_window - 1, .complete = TRUE
      ),
      innov_roll_sd = slider::slide_dbl(
        abs(innov), sd, .before = roll_sd_window - 1, .complete = TRUE
      ),
      innov_spike = abs(innov) > (innov_roll_mean + INNOV_MULT * innov_roll_sd)
    )
}


```



## 5. Exit Rules & Override

### 5.1 Exit Logic

+ Three exit conditions govern position closure. Take-profit triggers when the spread returns to the Kalman Filter mean estimate (z-score crosses zero). The time stop exits after `2 * half_life` days regardless of P&L, preventing capital from being trapped in broken pairs. The innovation override force-exits when the KF prediction errors spike for `max_innov_days` consecutive days, signaling a structural break in the pair relationship.


```{r}
#| label: exit-rules

apply_exit_rules <- function(signal_df, half_life_days) {
  # signal_df has z_score, innov_spike, and trade signals
  # Returns a tibble with final position accounting for exits

  time_stop_days <- max(round(2 * half_life_days), 5) # gives at least a trading week for each trade in case very short half-life

  signal_df %>%
    dplyr::mutate(
      # Raw signal from z-score
      raw_signal = dplyr::case_when(
        is.na(z_score) ~ 0L,
        z_score < -entry_thresh ~ 1L,    # spread negative > long pair
        z_score > entry_thresh  ~ -1L,   # spread positive > short pair
        TRUE ~ 0L
      ),
      innov_consec = slider::slide_dbl(
        as.numeric(innov_spike),
        sum, .before = N_INNOV_DAYS - 1, .complete = TRUE
      ),
      innov_override = dplyr::coalesce(innov_consec >= max_innov_days, FALSE)
    )
}

```


## 6. Monthly Rebalancing & Full Backtest

### 6.1 Backtesting

+ The backtest iterates month by month. At each month-end: Layer 1 re-clusters all 66 pairs and selects the tradeable set with inverse volatility parity weights. During the month, Layer 2 runs the Kalman Filter daily on each active pair. Pairs leaving the tradeable cluster are force-closed; pairs remaining are resized to new weights.


```{r}

backtest <- function(){
  
  # pull end of month dates for clustering rebalance
  
  # for first month of trading, pull last month's end date to have an initial cluster to    trade on
  
  # perform clustering at each month end - use function from earlier to choose tradeable    pairs
  
  # fit KF on tradeable pairs for each month and generate trade signals
    # KF has a warmup period so that we ensure it's initialization for the first month
      # is stable 
    # this must be done recursively, (for loop) because each day's trade decision depends       on last day's signal and trade
  
  # After signals are generated, add the safeguards:
    # z-score - if z score crosses 0 then exit because the spread mean-reverted
    # days_in_trade - if we hold a trade for more than 2*half_life that means it should've       # mean reverted by now so we exit regardless of price
  
    # Innovations - if innovations(errors) from KF fit are unusually large, that means          # that the relationship broke and a structural chg happened. So we exit to cut            # losses - could also mean we cut gains
  
  # To compute returns from trades, we assume that p1 is always long and p2 is always         short
    # The relationship is derived from the beta from the dynamic model fit from before
    # We weight each position by its inverse volatility (this ensures no single pair is
      # overweighted and that we dont cluster our risk)
    # if position yesterday was 1 (long) that means we are long p1 and short p2
    # if positions yesterday was -1 (short) that mean we short p1 and long p2
      # the return formula stays the same, the position sign filps it from long to short
  
  # Returns are calculated using yesterday's position * today's Cl-Cl return
  
  # All Trades/entry/exit are decided at close of each day, and executed at next open
  
  # get cum returns 
  
  
}


```


## 7. Parameter Optimization

### 7.1 Grid Search

## 8. Performance Results

### 8.1 Training Period Performance

### 8.2 Test Period Performance

### 8.3 RTL Trade Statistics

### Performance Visualized

## 9. Learnings